---
title: '1. Atlas PCa PDOS + tissue (tumor only) from Song, Hirz, Chen dataset: harmonization'
author: "Romuald Parmentier"
date: "2024-12-16"
output: html_document
---

# Prepare environement 

```{r Clone Github repo and/or point toward its direction}

library(git2r)

# Set the base directory where the repository should be cloned
base_dir <- "/Users/romuald/Bioinformatics/R_Projects"
setwd(base_dir)

# Load the git2r package (install if necessary)
if (!requireNamespace("git2r", quietly = TRUE)) {
  install.packages("git2r")
}

# Define the repository URL and the destination directory
repo_url <- "https://github.com/LeMagnenLab/Matrigel-free_PCa_PDOs.git"
github_dir <- file.path(base_dir, "Matrigel_Free_Github")

# Check if the repository already exists, clone if necessary
if (dir.exists(github_dir)) {
  message("Repository already exists at: ", github_dir)
} else {
  message("Cloning the repository into: ", github_dir)
  git2r::clone(url = repo_url, local_path = github_dir)
  message("Repository cloned successfully.")
  message("Github repository cloned in: ", github_dir)
}

```

```{r Get Zenodo file links}

library("rjson")
library("purrr")

# Load and read json fron zenodo
zenodo_json = list.files(github_dir, pattern = "json",full.names = T)
zenodo_json = fromJSON(paste0(readLines(zenodo_json), collapse = ""))

# Extract entries
zenodo_json_files <- zenodo_json$files$entries

# Convert to data.frame / tibble
zenodo_files_df <- tibble::tibble(
       key = names(zenodo_json_files),
       mimetype = map_chr(zenodo_json_files, ~ .x$mimetype),
       url = map_chr(zenodo_json_files, ~ .x$links$content),
       size = map_dbl(zenodo_json_files, ~ .x$size)
   )


```

```{r Load libraries and create output folder, message = F}

# Libraries
library(SingleCellExperiment)
library(dplyr)

# Functions and palettes
source(paste0(github_dir, "/Dolgos_Custom_Functions.R"))
source(paste0(github_dir, "/Dolgos_Custom_Color_Palettes.R"))

# Create output path
out_path = create_exp_folder(
  github_dir = github_dir,
  samples_ID = "Atlas_PCa_PDO-Song_Tissue-Hirz_Tissue-Chen_Tissue",
  exp = "1_harmonization"
)

```

# Load and harmonize all data from different dataset

## Load metadata of all samples included in this atlas

This csv file has been made manually and has been stored in the same folder

```{r Load metadata from atlas}

sce_metadata <- zenodo_files_df %>% 
  filter(key == "Atlas_PCa_PDOs_metadata.csv") %>% 
  pull(url) %>% 
  read.csv()

```


## Load PDO atlas to retrieve metadata for all PDO datasets

```{r Load sample metadata}

file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Atlas_PCa_PDOs",
  prev_exp = "8_final_export", 
  pattern = "Atlas_PCa_PDOs_sce_final.rds")

Dolgos_Atlas_PDO_sce_comb = readRDS(file_path)

```

### McCray_2019 PDO

```{r Load normalized sce}

## Load the only normalized sce for PDOs
## Annotation used for this sce is the one computed in Dologos_2024 with SingleR (stored in Dolgos_Atlas_PDO_sce_comb)

file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "McCray_PDOs",
  prev_exp = "2_normalization", 
  pattern = "sce_normalized.rds")

McCray_sce = readRDS(file_path)

```

```{r Structuring samples metadata}

# Filter for the correct sample
sce_metadata = samples_metadata %>%
  filter(Dataset == "McCray_2019")

# Creta the metadta df by replciating the metadata by he number of cells
sce_metadata = sce_metadata %>%
  slice(rep(1, ncol(McCray_sce))) 

# Add cell specific sizeFacor column 
sce_metadata$sizeFactor = McCray_sce$sizeFactor

# Store colnames
cell_names = colnames(McCray_sce)

# Replace the colData by the new table
colData(McCray_sce) = DataFrame(sce_metadata)
colnames(McCray_sce) = cell_names

# Subsetting unwanted cells in the sce
cells_in_sce =  which(colnames(Dolgos_Atlas_PDO_sce_comb) %in% colnames(McCray_sce))
McCray_sce$Cell_Type = Dolgos_Atlas_PDO_sce_comb$Cell_Type[cells_in_sce]

# Store the sce in a list for convenience in later steps
McCray_list_sce = list(McCray_sce)
names(McCray_list_sce) = unique(McCray_sce$Sample_Original_ID)

```

## Song_2022 PDO and Tissue

```{r Load list of normalized sce objects and annotated combined object}

## Load the list of normalized sce for PDOs 
## Annotation used for these sce is the one computed in Dologos_2024 with SingleR (stored in Dolgos_Atlas_PDO_sce_comb)
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Song_PDOs",
  prev_exp = "2_normalization_rescaling", 
  pattern = "list_sce_normalized.rds")

Song_Org_list_sce = readRDS(file_path)

## Load the list of normalized sce for tissue 
## Annotation used for these sce is the original one contained in the original object (Song_Tissue_sce_comb)
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Song_Tissue",
  prev_exp = "1bis_normalization_Rescaling",
  pattern = "list_sce_normalized.rds")

Song_Tissue_list_sce = readRDS(file_path)

# Load the annotated tissue object with original annotation (epithelial cells only)
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Song_Tissue",
  prev_exp = "1bis_normalization_Rescaling",
  pattern = "sce_comb_normalized_rescaled.rds")

Song_Tissue_sce_comb = readRDS(file_path)

```

```{r Structuring samples metadata for PDOs}

Song_Org_list_sce = lapply(X = Song_Org_list_sce, FUN = function(sce){
  
  # Filter for the correct sample
  sce_metadata = samples_metadata %>%
    filter(Dataset == "Song_2022" & Sample_Name == unique(sce$Sample_Description))
  
  # Creta the metadta df by replciating the metadata by he number of cells
  sce_metadata = sce_metadata %>%
    slice(rep(1, ncol(sce))) 
  
  # Add cell specific sizeFacor column 
  sce_metadata$sizeFactor = sce$sizeFactor
  
  # Store colnames
  cell_names = colnames(sce)
  
  # Replace the colData by the new table
  colData(sce) = DataFrame(sce_metadata)
  colnames(sce) = cell_names
  
  # Subsetting unwanted cells in the sce
  cells_in_sce =  which(colnames(Dolgos_Atlas_PDO_sce_comb) %in% colnames(sce))
  sce$Cell_Type = Dolgos_Atlas_PDO_sce_comb$Cell_Type[cells_in_sce]
  
  return(sce)
  
} )

```

```{r Structuring samples metadata for tissue}

Song_Tissue_list_sce = lapply(X = Song_Tissue_list_sce, FUN = function(sce){
  
  # Filter for the correct sample
  sce_metadata = samples_metadata %>%
    filter(Dataset == "Song_2022" & Sample_Name == unique(sce$Sample_Description))
  
  # Creta the metadta df by replciating the metadata by he number of cells
  sce_metadata = sce_metadata %>%
    slice(rep(1, ncol(sce))) 
  
  # Add cell specific sizeFacor column 
  sce_metadata$sizeFactor = sce$sizeFactor
  
  # Store colnames
  cell_names = colnames(sce)
  
  # Replace the colData by the new table
  colData(sce) = DataFrame(sce_metadata)
  colnames(sce) = cell_names
  
  # Subsetting unwanted cells in the sce
  cells_in_sce =  which(colnames(Song_Tissue_sce_comb) %in% colnames(sce))
  sce$Cell_Type = Song_Tissue_sce_comb$ID[cells_in_sce]
  
  return(sce)
  
} )

```

# Chen_2022 tissue

```{r Load list of normalized sce objects and annotated combined object}

## Load the list of normalized sce  
## Annotation used for these sce is the original one contained in the original object (Chen_sce_comb)
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Chen_Tissue",
  prev_exp = "2_normalization_rescaling", 
  pattern = "list_sce_normalized.rds")

Chen_list_sce = readRDS(file_path)

# Load the annotated tissue object with original annotation (all cell types)
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Chen_Tissue",
  prev_exp = "4bis_dimension_reduction_all_cells", 
  pattern = "sce_comb_dim_reduced")

Chen_sce_comb = readRDS(file_path)

```

```{r Fix colnames}

# Apply the function to each SingleCellExperiment in the list
Chen_list_sce <- lapply(Chen_list_sce, function(sce) {
    
  # Assign new names
  colnames(sce) = sub("_", "-", colnames(sce))
  
  # Update sce in the list
  return(sce)  # Return the modified SingleCellExperiment
  
})

# change colnames of the sce_conb object
colnames(Chen_sce_comb) = sub("_", "-", colnames(Chen_sce_comb))


```

```{r Structuring samples metadata}

Chen_list_sce = lapply(X = Chen_list_sce, FUN = function(sce){
  
  # Subsetting unwanted cells in the sce
  names_epithelial_cells = colnames(Chen_sce_comb)[which(Chen_sce_comb$broad_type == "Epithelial cells")]
  
  sce = sce[, which(colnames(sce) %in% names_epithelial_cells)]
  
  # Filter for the correct sample
  sce_metadata = samples_metadata %>%
    filter(Dataset == "Chen_2022" & Sample_Name == unique(sce$Sample_Description))
  
  # Creta the metadta df by replciating the metadata by he number of cells
  sce_metadata = sce_metadata %>%
    slice(rep(1, ncol(sce))) 
  
  # Add cell specific sizeFacor column 
  sce_metadata$sizeFactor = sce$sizeFactor
  
  # Store colnames
  cell_names = colnames(sce)
  
  # Replace the colData by the new table
  colData(sce) = DataFrame(sce_metadata)
  colnames(sce) = cell_names
  
  # Subsetting unwanted cells in the sce
  cells_in_sce =  which(colnames(Chen_sce_comb) %in% colnames(sce))
  sce$Cell_Type = Chen_sce_comb$broad_type[cells_in_sce]
  
  return(sce)
  
})


```

## Huang_2023 PDO

```{r Load list of normalized sce objects with original annotation}

## Load the list of normalized sce
## Annotation used for this sce is the one computed in Dologos_2024 with SingleR (stored in Dolgos_Atlas_PDO_sce_comb)
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Huang_PDOs",
  prev_exp = "1_normalization_rescaling", 
  pattern = "list_sce_normalized.rds")

Huang_list_sce = readRDS(file_path)

```

```{r Structuring samples metadata}

Huang_list_sce = lapply(X = Huang_list_sce, FUN = function(sce){
  
  # Subsetting unwanted cells in the sce
  untreated_cells = which(sce$Treatment == "DHT")
  sce = sce[,untreated_cells]
  
  # Filter for the correct sample
  sce_metadata = samples_metadata %>%
    filter(Dataset == "Huang_2023" & Sample_Name == unique(sce$Sample_ID))
  
  # Creta the metadta df by replciating the metadata by he number of cells
  sce_metadata = sce_metadata %>%
    slice(rep(1, ncol(sce))) 
  
  # Add cell specific sizeFacor column 
  sce_metadata$sizeFactor = sce$sizeFactor
  
  # Store colnames
  cell_names = colnames(sce)
  
  # Replace the colData by the new table
  colData(sce) = DataFrame(sce_metadata)
  colnames(sce) = cell_names
  
  # Subsetting unwanted cells in the sce
  cells_in_sce =  which(colnames(Dolgos_Atlas_PDO_sce_comb) %in% colnames(sce))
  sce$Cell_Type = Dolgos_Atlas_PDO_sce_comb$Cell_Type[cells_in_sce]
  
  return(sce)
  
})

```

# Hirz_2023 tissue

```{r Load list of normalized sce objects and combined object with original annotation}

## Load the list of normalized sce  
## Annotation used for these sce is the original one contained in the original object (Hirz_sce_comb)
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Hirz_Tissue",
  prev_exp = "2_normalization_rescaling",
  pattern = "list_sce_normalized.rds")

Hirz_list_sce = readRDS(file_path)

# Load the annotated tissue object with original annotation (all cell types)
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Hirz_Tissue",
  prev_exp = "3_dimension_reduction",
  pattern = "sce_comb_dim_reduced")

Hirz_sce_comb = readRDS(file_path)

```

```{r Fix colnames}

# Apply the function to each SingleCellExperiment in the list
Hirz_list_sce <- lapply(Hirz_list_sce, function(sce) {
  
  # Make new names by reordering prex and sufix from previous names
  new_names = sapply(colnames(sce), function(name) {
    parts <- strsplit(name, "_")[[1]]  # Split the string by "_"
      suffix <- parts[1]
      prefix <- sub("-1$", "", parts[2])  # Remove the trailing "-1"
      paste0(prefix, "-", suffix)  # Concatenate in the desired order
  })
    
  # Assign new names
  colnames(sce) = new_names
  
  # Update sce in the list
  return(sce)  # Return the modified SingleCellExperiment
  
})


```

```{r Structuring samples metadata}

Hirz_list_sce = lapply(X = Hirz_list_sce, FUN = function(sce){
  
  # Subsetting unwanted cells in the sce
  names_epithelial_cells = colnames(Hirz_sce_comb)[grepl(x = Hirz_sce_comb$cell_type_broad,pattern = "Epithelial")]
  
  sce = sce[, which(colnames(sce) %in% names_epithelial_cells)]
  
  # Filter for the correct sample
  sce_metadata = samples_metadata %>%
    filter(Dataset == "Hirz_2023" & Sample_Name == unique(sce$Sample_ID))
  
  # Creta the metadta df by replciating the metadata by he number of cells
  sce_metadata = sce_metadata %>%
    slice(rep(1, ncol(sce))) 
  
  # Add cell specific sizeFacor column 
  sce_metadata$sizeFactor = sce$sizeFactor
  
  # Store colnames
  cell_names = colnames(sce)
  
  # Replace the colData by the new table
  colData(sce) = DataFrame(sce_metadata)
  colnames(sce) = cell_names
  
  # Subsetting unwanted cells in the sce
  cells_in_sce =  which(colnames(Hirz_sce_comb) %in% colnames(sce))
  sce$Cell_Type = Hirz_sce_comb$cell_type_broad[cells_in_sce]
  
  return(sce)
  
})


```

# Dolgos_2024 PDO & tissue

```{r Load list of normalized sce objects and combined object with original annotation}

## Load the list of normalized sce (PDO and tissue)
## Annotation used for this sce is the one computed in Dologos_2024 with SingleR (stored in Dolgos_Atlas_PDO_sce_comb)
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Dolgos_PDOs_All-Dolgos_Tissue_All",
  prev_exp = "normalization_rescaling", 
  pattern = "list_sce_normalized.rds")

Dolgos_list_sce = readRDS(file_path)

# Load the annotated object of the PDO atlas to retrieve cell type of the organoids
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Atlas_PCa_PDOs",
  prev_exp = "4_cell_type_annotation", 
  pattern = "sce_comb_cell_type_annotated.rds")

Dolgos_Atlas_PDO_sce_comb = readRDS(file_path)

# Load the annotated object of the matched Tissue-PDO to retrieve cell type of the tissue
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Dolgos_PDOs-Dolgos_Tissue_Matching",
  prev_exp = "4_malignancy_annotation", 
  pattern = "sce_comb_malignancy_annotated.rds")

Dolgos_Tissue_sce_comb = readRDS(file_path)


```

```{r Structuring samples metadata}

Dolgos_list_sce = lapply(X = Dolgos_list_sce, FUN = function(sce){
  
  # Get sample name
  sample_name = unique(sce$Sample_Description)
  
  print(sample_name)
  
  # Filter for the correct sample
  sce_metadata = samples_metadata %>%
    filter(Dataset == "Dolgos_2024" & Sample_Name == unique(sce$Sample_Description))
  
  # Creta the metadta df by replciating the metadata by he number of cells
  sce_metadata = sce_metadata %>%
    slice(rep(1, ncol(sce))) 
  
  # Add cell specific sizeFacor column 
  sce_metadata$sizeFactor = sce$sizeFactor
  
  # Store colnames
  cell_names = colnames(sce)
  
  # Replace the colData by the new table
  colData(sce) = DataFrame(sce_metadata)
  colnames(sce) = cell_names
  
  # Subsetting unwanted cells in the sce
  if(grepl(x = sample_name, pattern = "Mat|ECMf")){ # If sqample is a PDO get the cell type from the TPDO Atlas object
    cells_in_sce =  which(colnames(Dolgos_Atlas_PDO_sce_comb) %in% colnames(sce))
    sce$Cell_Type = Dolgos_Atlas_PDO_sce_comb$Cell_Type[cells_in_sce]
  }else{ # If sample is a Tissue get the cell type from the Tissue sce object
    cells_in_sce =  which(colnames(Dolgos_Tissue_sce_comb) %in% colnames(sce))
    sce$Cell_Type = Dolgos_Tissue_sce_comb$Cell_Type[cells_in_sce]
  }
  
  return(sce)
  
})


```

# Merge all list of sce

```{r Merging all list of sce and cleaning environement}

# Gather all sce_comb in a single list
list_all_sce = list(
  Huang_list_sce, McCray_list_sce, Song_Org_list_sce, Dolgos_list_sce, 
  Chen_list_sce, Song_Tissue_list_sce,Hirz_list_sce)

list_all_sce <- do.call(c, list_all_sce)

rm(
  Chen_sce_comb, Dolgos_Atlas_PDO_sce_comb, McCray_sce , Song_Tissue_sce_comb, Hirz_sce_comb, Dolgos_Tissue_sce_comb,
  Huang_list_sce, Chen_list_sce, Dolgos_list_sce, Hirz_list_sce, McCray_list_sce, Song_Org_list_sce, Song_Tissue_list_sce, Dolgos_list_sce
  )

gc()

```

# Harmonizing col/rownames col/rowData

```{r Harmonizing count matrix rownames and colnames}

##############################
### rownames (= features) ###
#############################

list_all_sce_restricted = restrict_reorder_features_names(sce_list = list_all_sce)

# list_all_sce_extended = extend_and_reorder_SCE(sce_list = list_all_sce)

##################################
### colnames (= cell barcodes) ###
##################################

# Check colnames of the matrix (normally done in previous steps)
lapply(list_all_sce_restricted, function(x){head(colnames(x))})

```

```{r Harmonizing rowData and coldata}

###############
### RowData ###
###############

# Identify the common rowData columns across all SCE objects
common_columns <- Reduce(intersect, lapply(list_all_sce_restricted, function(sce) colnames(rowData(sce))))

# Update each SCE object to retain only the common rowData columns
list_all_sce_restricted <- lapply(list_all_sce_restricted, function(sce) {
  rowData(sce) <- rowData(sce)[, common_columns, drop = FALSE]  # Subset rowData
  return(sce)  # Return updated sce
})

# Check colnames of the rowData (performed above)
lapply(list_all_sce_restricted, function(x){colnames(rowData(x))})

###############
## ColData ###
##############

# Check colnames of the colData (normally done in previous steps)
lapply(list_all_sce_restricted, function(x){colnames(colData(x))})

```

# Create a singkle sce_comb object from the list of sce

```{r Creating a sce_comb object with all sce}

## Get total raw counts, first in a list, then combine them into a matrix
counts_comb = lapply(list_all_sce_restricted, function(x) {counts(x)})
counts_comb <- Reduce(function(x, y) cbind(x, y), counts_comb)

## Then get total logNormed counts, first in a list, then combine them into a matrix
logcounts_comb = lapply(list_all_sce_restricted, function(x) {logcounts(x)})
logcounts_comb <- Reduce(function(x, y) cbind(x, y), logcounts_comb)

# Combine all cell metadata (columns are already the same for each object of the list)
colData_comb = lapply(list_all_sce_restricted, function(x) {colData(x)})
colData_comb = Reduce(function(x, y) rbind(x, y), colData_comb)

# Combine the slots to create the combined sce object
sce_comb <- SingleCellExperiment( 
  assays = list(counts = counts_comb, logcounts = logcounts_comb),  
  rowData = rowData(list_all_sce_restricted[[1]]), # RowData is similar among all elements 
  colData = colData_comb
) 

```

```{r Export list_sce and sce_comb}

saveRDS(
  object = list_all_sce_restricted,
  file = paste0(out_path,time_stamp(),"list_all_sce_restricted_comb_harmonized.rds"))

saveRDS(
  object = sce_comb,
  file = paste0(out_path,time_stamp(),"sce_comb_all_restricted_harmonized.rds"))

```

