---
title: "3.Chen 2021 (Tumor Tissue): broad cell type annotation"
author: "Romuald Parmentier"
date: "2024-10-28"
output: html_document
---

```{r Load libraries and create output folder, message = F}

# Libraries
library(Seurat)
library(SingleCellExperiment)
library(ggplot2)
library(dplyr)
library(SingleR) # Cell type annotation
library(celldex)

# Functions and palettes
source(paste0(github_dir, "/Dolgos_Custom_Functions.R"))
source(paste0(github_dir, "/Dolgos_Custom_Color_Palettes.R"))

# Create output path
out_path = create_exp_folder(
  github_dir = github_dir,
  samples_ID = "Chen_Tissue",
  exp = "3_broad_cell_type_annotation"
)


```

# Prepare the data

```{r Load the files}

# Load rescaled atlas sce object after normalization and rescaling
file_path = get_exp_file_path(
  github_dir = github_dir,
  samples_ID = "Chen_Tissue",
  prev_exp = "2_normalization_rescaling", 
  pattern = "sce_comb_normalized_rescaled.rds")

sce_comb = readRDS(file = file_path) 

# Load BlueprintEncodeData ref for broad cell type assignation
file_path = "/scicore/home/wykopa75/GROUP/rparmentier/sc_RNAseq/Projects/data/BlueprintEncodeData/BlueprintEncodeData.rds"

Blueprint_ref = readRDS(file = file_path) 

# Load celltype_ref_sce object from Hirz 2022
####################################

file_path = get_exp_file_path(
  organ = "Prostate",
  project = "Medium_Optimization",
  samples_ID = "Hirz_2022",
  prev_exp = "annotate_harmonize_normalize_sce", 
  pattern = "sce_comb_QuickClustered_MultiBatch_normed.rds")

print(file_path)
hirz_ref = readRDS(file = file_path) 

```

# Broad cell type annotation

```{r}

# Compute the prediction
pred_type <- SingleR(
  test = sce_comb, 
  ref = Blueprint_ref, 
  labels = Blueprint_ref$label.main)

# Assign braod cell types to tissues
sce_comb$broad_type_Blueprint = pred_type$pruned.labels

table(sce_comb$Sample_Description, sce_comb$broad_type_Blueprint)

```

# Annotation according Hirz_2022 dataset

```{r}

hirz_ref = hirz_ref[, sample(ncol(hirz_ref), size = ncol(sce_comb), replace = FALSE)]
gc()

pred_broad <- SingleR(
  test = sce_comb, 
  ref = hirz_ref, 
  labels = hirz_ref$cell_type_broad)

# Adding the cell type is done on the origianl sce object so that the genes removed during the restrcit rowDt step are back
sce_comb$broad_type_Hirz = pred_broad$pruned.labels

```


# Visualization 

No visualization here as the dimension redution has not been performed yet. 
==> This is was wanted as we wanted to perform dimension reduction in parallel for both epithelial only and all cell type. Thus cell type assignation had to be perfroned beforehand.

# Export files

```{r Export sce_comb object}

saveRDS(
  object = sce_comb,
  file = paste0(out_path,time_stamp(),"sce_comb_broad_cell_type_annotated.rds"))

```
